\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepackage{hyperref}
\usepackage{xurl}
\usepackage{microtype}
\emergencystretch=3em

\newtheorem{theorem}{Theorem}
\newtheorem{conjecture}{Conjecture}
\newtheorem{proposition}{Proposition}

\newcommand{\degmin}{\delta}

\title{A SAT-Modulo-Symmetries verification of the Erd\H{o}s--Gy\'arf\'as
power-of-two cycle conjecture for minimum-degree-3 graphs up to 30 vertices}

\author{Arjun Balaji\thanks{Columbia University. ORCID:
\href{https://orcid.org/0009-0005-1790-0034}{0009-0005-1790-0034}. This is a
computational result; see \S\ref{sec:verification} for the verification and
\S\ref{sec:limitations} for scope.} \\ Columbia University}

\date{\today}

\begin{document}
\maketitle

\begin{abstract}
The Erd\H{o}s--Gy\'arf\'as conjecture (1995) asserts that every graph of minimum
degree at least $3$ contains a cycle whose length is a power of two. It is open.
Prior computer search established that any general minimum-degree-3 counterexample
has at least $17$ vertices (Royle and Markstr\"om), and any cubic (3-regular)
counterexample at least $30$ (Markstr\"om, 2004). We give, to our knowledge, the
first application of SAT-based methods to this conjecture. Using SAT Modulo
Symmetries with the Glasgow subgraph solver as a complete forbidden-subgraph
propagator, we verify that every minimum-degree-3 graph on at most $30$ vertices
contains a cycle of length $4$, $8$, or $16$. Hence any general minimum-degree-3
counterexample has at least $31$ vertices, improving the two-decade-old general
bound from $17$ to $31$; since the cubic class is contained in this class, the cubic
bound improves from $30$ to $31$ as well. We corroborate the result with an exact
ground-truth check at $n=10$, reproduction of the $n\le 16$ baseline, agreement with
an independent CEGAR-SAT solver for $n\le 19$, and robustness across cardinality
encodings and symmetry-breaking methods, and we discuss the path to a
machine-checked proof certificate.
\end{abstract}

\medskip
\noindent\textbf{Keywords:} Erd\H{o}s--Gy\'arf\'as conjecture; power-of-two cycle;
minimum degree; SAT modulo symmetries; forbidden subgraph; computer-assisted
enumeration.

\smallskip
\noindent\textbf{2020 Mathematics Subject Classification:}
05C38 (primary); 05C30, 05C85, 68V05 (secondary).

\section{Introduction}
\label{sec:intro}

In 1995 Erd\H{o}s and Gy\'arf\'as posed the following conjecture (see
Erd\H{o}s~\cite{erdos1997}).

\begin{conjecture}[Erd\H{o}s--Gy\'arf\'as]
\label{conj:eg}
Every graph with minimum degree at least $3$ contains a simple cycle whose
length is a power of two.
\end{conjecture}

Here ``power of two'' means $2^k$ for some integer $k \ge 2$, i.e.\ a cycle of
length $4, 8, 16, \dots$. The conjecture remains open; Erd\H{o}s offered \$100 for
a proof and \$50 for a counterexample~\cite{erdos-problems}. It has been confirmed
for several restricted classes, including 3-connected cubic planar
graphs~\cite{heckman-krakovski}, $P_8$-free graphs~\cite{gao-shan}, and $P_{10}$-free
graphs~\cite{hu-shen}; the most recent structural progress shows that any
\emph{minimal} counterexample is ``predominantly cubic''~\cite{carr2026}.

On the computational side, two distinct frontiers are known, and the distinction
is central to the present paper. Write $\degmin(G)$ for the minimum degree of $G$.
\begin{itemize}
  \item \textbf{General} ($\degmin \ge 3$, arbitrary maximum degree): computer
  searches of Royle and Markstr\"om are commonly cited as establishing that any
  counterexample has at least $17$ vertices, i.e.\ Conjecture~\ref{conj:eg} holds
  for all $\degmin\ge 3$ graphs on at most $16$ vertices~\cite{wikipedia-eg}.
  \item \textbf{Cubic} (3-regular): Markstr\"om~\cite{markstrom2004} verified all
  cubic graphs on at most $29$ vertices, so any cubic counterexample has at least
  $30$ vertices.
\end{itemize}
A targeted, adversarial literature search (covering theses, technical reports,
OEIS and House of Graphs, and the two most recent surveys of computer-assisted
graph theory and of SAT Modulo Symmetries~\cite{szeider-survey,jooken-survey})
found no improvement of the \emph{general} bound past $17$ after 2004, and no prior
application of SAT, SAT Modulo Symmetries, constraint programming, or isomorph-free
generation to Conjecture~\ref{conj:eg}.

\paragraph{Contribution.}
We give the first SAT-based attack on Conjecture~\ref{conj:eg}. Using SAT Modulo
Symmetries (SMS)~\cite{sms-tocl,szeider-survey} together with the Glasgow subgraph
solver~\cite{glasgow}, we verify Conjecture~\ref{conj:eg} for \emph{all}
minimum-degree-3 graphs on $n$ vertices for every $n$ up to $30$:

\begin{theorem}
\label{thm:main}
Every graph with minimum degree at least $3$ on at most $30$ vertices contains a
cycle of length $4$, $8$, or $16$. Consequently, any minimum-degree-3
counterexample to the Erd\H{o}s--Gy\'arf\'as conjecture has at least $31$ vertices.
\end{theorem}

This improves the general bound from $17$ to $31$, a $+14$-vertex extension over a
search space that, unlike the cubic case, includes all non-regular graphs of
minimum degree $3$.

\paragraph{General versus cubic (a point to read carefully).} Our verified order,
$30$, coincides numerically with Markstr\"om's cubic \emph{bound} of $30$, but the
two statements are different and should not be conflated. Markstr\"om verified the
\emph{cubic} class up to order $29$ (cubic bound $\ge 30$). We verify the
\emph{entire} minimum-degree-3 class (a strictly larger family, dominated by
non-regular graphs) up to order $30$ (general bound $\ge 31$). Since the cubic
class is contained in the minimum-degree-3 class, our search in particular
verifies all cubic graphs on $\le 30$ vertices, which incidentally also improves
the cubic bound from $30$ to $31$. Thus the genuinely new content is twofold: the
\emph{non-cubic} minimum-degree-3 graphs on $17 \le n \le 30$ vertices, and the
methodology (SAT/SMS), which had not previously been applied to this conjecture.

\paragraph{Related work.} The conjecture is known for several restricted classes:
$K_{1,m}$-free graphs~\cite{shauger}, planar claw-free graphs~\cite{daniel-shauger},
3-connected cubic planar graphs~\cite{heckman-krakovski}, and the $P_t$-free
families $P_8$~\cite{gao-shan}, $P_{10}$~\cite{hu-shen}, and (with the aid of a
computer search) $P_{13}$~\cite{hegde2024}. These are theorems for infinite
hereditary classes and do not bound the order of a general counterexample; in
particular, the computer search of~\cite{hegde2024} assists a structural proof for
the $P_{13}$-free class and is unrelated to the exhaustive, isomorph-free
SAT enumeration used here. The
most recent structural progress, Carr~\cite{carr2026}, shows that at least $4/7$ of
the vertices of any minimal counterexample have degree exactly $3$; it is an
arXiv preprint that uses no computation. Our methodological ingredients, SAT
Modulo Symmetries~\cite{sms-tocl,szeider-survey} and the Glasgow subgraph
solver~\cite{glasgow}, have been applied to other graph-existence questions
(e.g.\ the 3-Decomposition Conjecture on cubic graphs), but, as noted above, not
to Conjecture~\ref{conj:eg}.

\section{Method}
\label{sec:method}

For a fixed number of vertices $n$ we decide the following question with a single
SMS call: \emph{does there exist a graph $G$ on $n$ vertices with $\degmin(G)\ge 3$
that contains no cycle of length $4$, $8$, or $16$?} The reduction to these three
cycle lengths is justified by the following elementary observation.

\begin{proposition}
\label{prop:reduction}
For $3 \le n \le 31$, a graph on $n$ vertices contains a cycle whose length is a
power of two if and only if it contains $C_4$, $C_8$, or $C_{16}$ as a
(non-induced) subgraph.
\end{proposition}
\begin{proof}
A cycle of length $\ell$ is precisely a $C_\ell$ subgraph. The powers of two that
are valid cycle lengths (at least $3$) and at most $n \le 31$ are exactly $4, 8,
16$; the next one, $32$, exceeds $n$.
\end{proof}

By Proposition~\ref{prop:reduction}, if for every $n \le 30$ no minimum-degree-3
graph on $n$ vertices contains $C_4$, $C_8$, or $C_{16}$ is found, then no
minimum-degree-3 counterexample on at most $30$ vertices exists. A minimum-degree-3
graph requires $n \ge 4$, and for $n \le 16$ the conjecture is the established
baseline, which we reproduce ($\S\ref{sec:verification}$; the smallest orders $n=4,5$
are also covered by the independent CEGAR check below, and hold trivially since
$K_4$ contains $C_4$). It therefore suffices to settle $17 \le n \le 30$, which is the
content of Theorem~\ref{thm:main}.

\paragraph{SAT Modulo Symmetries.} SMS performs complete, isomorph-free
generation of graphs satisfying given constraints: its modified CaDiCaL solver
carries an in-search canonicity (minimality) propagator that prunes any partial
adjacency matrix that cannot be the lexicographically minimal representative of
its isomorphism class~\cite{sms-tocl}. This is the key to scaling: the search
considers essentially one graph per isomorphism class rather than enumerating all
labelings.

\paragraph{Encoding.} The minimum-degree constraint $\degmin(G) \ge 3$ is encoded
as a CNF cardinality constraint over the $\binom{n}{2}$ edge variables (we use
SMS's built-in sequential-counter encoding; see \S\ref{sec:verification} for a
totalizer cross-check). The constraint ``no $C_4$, $C_8$, or $C_{16}$'' is enforced
by the Glasgow subgraph solver~\cite{glasgow} as a complete forbidden-subgraph
propagator: whenever a (partial) graph contains one of these cycles as a
(non-induced) subgraph, the propagator adds a clause excluding it. Forbidding the
short cycle $C_4$ statically in CNF is cheap, but $C_8$ and $C_{16}$ have
$\Theta(n^8)$ and $\Theta(n^{16})$ potential occurrences, so a static encoding is
infeasible; the propagator is what makes the search tractable.

\paragraph{Why not plain CEGAR-SAT.} As a baseline we also implemented a
self-contained counterexample-guided (CEGAR) SAT solver in Python (PySAT with
CaDiCaL, a DFS power-of-two cycle detector, and a static adjacent-transposition
lexicographic symmetry break). With only a partial symmetry break it re-discovers
each forbidden cycle in many isomorphic copies: at $n=17$ it accumulates roughly
$85{,}000$ refinement clauses and the per-instance time grows by about a factor of
two per vertex, so it stalls at $n=19$ (bound $\ge 20$). SMS decides the same
$n=17$ instance in $2.9$ seconds (Table~\ref{tab:results}). Complete symmetry
breaking is precisely what removes this bottleneck; the two solvers agree wherever
both finish ($n \le 19$, \S\ref{sec:verification}).

\section{Results}
\label{sec:results}

Table~\ref{tab:results} reports, for each $n$ from $17$ to $30$, that SMS finds
\emph{no} minimum-degree-3 graph avoiding all of $C_4, C_8, C_{16}$ (solution count
$0$). All runs used a single CPU core.

\begin{table}[h]
\centering
\begin{tabular}{rrr}
\toprule
$n$ & count & wall time (s) \\
\midrule
17 & 0 & 2.9 \\
18 & 0 & 7.8 \\
19 & 0 & 24.1 \\
20 & 0 & 27.8 \\
21 & 0 & 19.3 \\
22 & 0 & 101.1 \\
23 & 0 & 202.9 \\
24 & 0 & 148.3 \\
25 & 0 & 339.8 \\
26 & 0 & 414.8 \\
27 & 0 & 1000.3 \\
28 & 0 & 1892.0 \\
29 & 0 & 2342.8 \\
30 & 0 & 6888.9 \\
\bottomrule
\end{tabular}
\caption{SMS decision per order $n$: no minimum-degree-3 graph on $n$ vertices
avoids $C_4, C_8, C_{16}$. Together with the reproduced $n\le16$ baseline this
proves Theorem~\ref{thm:main}.}
\label{tab:results}
\end{table}

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{semilogyaxis}[
  width=0.82\linewidth, height=5.6cm,
  xlabel={order $n$}, ylabel={wall time (s)},
  xtick={17,19,21,23,25,27,29}, ymin=1, ymax=15000,
  grid=both, mark size=1.4pt,
]
\addplot[mark=*] coordinates {
(17,2.9)(18,7.8)(19,24.1)(20,27.8)(21,19.3)(22,101.1)(23,202.9)(24,148.3)
(25,339.8)(26,414.8)(27,1000.3)(28,1892.0)(29,2342.8)(30,6888.9)
};
\end{semilogyaxis}
\end{tikzpicture}
\caption{Single-core wall time of the SMS decision at each order $n$ (log scale).
Every point is UNSAT (no minimum-degree-3 graph on $n$ vertices avoids
$C_4, C_8, C_{16}$). Growth is roughly geometric, about a factor of $1.5$--$2$ per
added vertex.}
\label{fig:runtime}
\end{figure}

\paragraph{Experimental setup.} For each $n$ we encode minimum degree at least $3$
as a cardinality constraint over the $\binom{n}{2}$ edge variables (the
sequential-counter encoding produced by PySMS), and decide it with the SMS
solver~\cite{sms-tocl} run in enumeration mode, using the Glasgow subgraph
solver~\cite{glasgow} to forbid $C_4$, $C_8$, and $C_{16}$; a return of ``no
graphs'' certifies that no such graph exists. Each order was solved on a single CPU
core, and the wall times in Table~\ref{tab:results} exclude container start-up. The
exact solver invocation, the pinned tool versions, and the build and run scripts
are available in the accompanying repository.

\section{Verification}
\label{sec:verification}

A non-existence claim at $n=30$ cannot be brute-forced (exhaustive generation of
minimum-degree-3 graphs is infeasible beyond roughly $n=13$), so we corroborate
Theorem~\ref{thm:main} with several independent checks. \emph{No check produced a
contradiction.}
\begin{enumerate}
  \item \textbf{Ground-truth anchor.} At $n=10$, forbidding only $C_4$, SMS returns
  exactly $5$ graphs, matching the independent count of the $5$ $C_4$-free
  minimum-degree-3 graphs on $10$ vertices obtained with \texttt{nauty}
  (\texttt{geng}+\texttt{labelg})~\cite{nauty}.
  \item \textbf{Baseline reproduction.} For $n = 6, \dots, 16$, forbidding all of
  $C_4, C_8, C_{16}$ yields $0$, reproducing the published baseline.
  \item \textbf{Independent second solver.} Our CEGAR-SAT solver and SMS both
  return UNSAT for $n = 17, 18, 19$ (different solvers, symmetry breaking, and
  cycle handling).
  \item \textbf{Encoding robustness.} Re-deciding with the totalizer cardinality
  encoding (a structurally different CNF) gives count $0$ at $n = 17, 20, 22, 25$.
  \item \textbf{Symmetry-method robustness.} Re-deciding with SMS's colex
  minimality ordering gives count $0$ at $n = 17, 20$; the (slower, experimental)
  colex variant timed out at $n = 22, 25$ (inconclusive, not contradictory).
  \item \textbf{Positive controls.} Forbidding only $C_4$ (a weaker constraint)
  yields a graph at $n = 17, 20, 25, 30$, confirming the pipeline returns a
  solution when one exists.
\end{enumerate}

\paragraph{Toward a machine-checked certificate.} \texttt{smsg} can emit an LRAT
proof for UNSAT instances, but a generic LRAT checker cannot validate it against
the minimum-degree-3 CNF alone, because the forbidden-cycle clauses are added by
the propagator during search and are not RUP/RAT-derivable from that CNF (they
exclude valid minimum-degree-3 graphs). A fully machine-checked certificate
therefore requires the certified-SMS proof-logging machinery, in which the
propagator justifies its own clauses; we leave this as the principal next step.

\section{Discussion and limitations}
\label{sec:limitations}

Theorem~\ref{thm:main} advances the general minimum-degree-3 frontier from $16$ to
$30$ vertices (bound $17 \to 31$) and, as a byproduct, the cubic frontier from
$29$ to $30$ (bound $30 \to 31$). Two caveats apply. First, this is a
\emph{computational} result: its soundness rests on SMS's isomorph-free generation,
the completeness of the Glasgow subgraph propagator, and the correctness of the
min-degree encoding. The checks of \S\ref{sec:verification} corroborate this
composition (exactly, against \texttt{nauty} ground truth at $n=10$, and against
the established baseline for $n \le 16$), as is standard for computer-assisted
enumeration; the natural further strengthening, for a nonexistence result of this
kind, is an end-to-end machine-checked proof certificate (\S\ref{sec:conclusion}).
Second, the result does not prove the conjecture; it only raises the lower bound on
the order of a hypothetical counterexample.

\section{Conclusion and future work}
\label{sec:conclusion}

We have given the first SAT-based attack on the Erd\H{o}s--Gy\'arf\'as conjecture.
Casting the question as ``does a minimum-degree-3 graph with no $C_4$, $C_8$, or
$C_{16}$ exist?'' and discharging it with SAT Modulo Symmetries and the Glasgow
subgraph propagator, we verified the conjecture for all minimum-degree-3 graphs on
at most $30$ vertices, raising the general lower bound on a counterexample from
$17$ to $31$ (Theorem~\ref{thm:main}). The decisive ingredient is complete,
in-search symmetry breaking: a conventional CEGAR encoding stalls near $n=20$,
whereas SMS reaches $n=30$ in under two hours of single-core time.

Several directions remain. (i) \emph{Push the frontier.} The runtimes
(Figure~\ref{fig:runtime}) suggest $n=31$ and a few further orders are within reach
of a longer single run or modest parallelism; the present search was halted at
$n=30$. (ii) \emph{A machine-checked certificate.} Replacing the verification
checks of \S\ref{sec:verification} by an end-to-end, formally verified proof (via
the certified-SMS proof-logging machinery, so that the forbidden-subgraph and
canonicity propagators justify their own clauses) would put the result on the
same footing as certified SAT resolutions of, e.g., the Pythagorean triples and
Schur-number problems. (iii) \emph{Exploit structure.} Carr's result that a minimal
counterexample is predominantly cubic~\cite{carr2026} could be added as a
propagator to prune the search and reach larger $n$. (iv) \emph{Other questions.}
The same SMS-plus-forbidden-subgraph template applies directly to related
cycle-spectrum problems. An independent reproduction, or the proof certificate of
(ii), would further raise confidence in the result.

\section*{Code and data availability}
All source code, the per-order data, the exact build script (with pinned tool
commits), and the verification scripts are openly available, enabling independent
reproduction of every value in Table~\ref{tab:results}, at
\url{https://github.com/ArjunBalaji79/erdos-gyarfas-min-degree-3}.

\section*{Disclosure statement}
The author reports no competing interests.

\section*{Disclosure of generative AI use}
In preparing this work the author used Anthropic's Claude (the Claude~Opus~4.8
model, via the Claude~Code interface) to implement the SAT/SMS pipeline and the
verification scripts, to configure and run the computational experiments, and to
assist with drafting and the literature review. The author reviewed the generated
code, outputs, and text, checked the results through the verification of
\S\ref{sec:verification}, and takes full responsibility for the content and claims
of this paper. No AI system is an author.

\begin{thebibliography}{99}
\bibitem{erdos1997} P.~Erd\H{o}s, \emph{Some old and new problems in various
branches of combinatorics}, Discrete Math.\ 165/166 (1997) 227--231.
\bibitem{erdos-problems} Erd\H{o}s Problems, ``Power-of-two cycles'',
\url{https://www.erdosproblems.com/}.
\bibitem{markstrom2004} K.~Markstr\"om, \emph{Extremal graphs for some problems on
cycles in graphs}, Congressus Numerantium 171 (2004) 177--188.
\bibitem{heckman-krakovski} C.~C.~Heckman, R.~Krakovski, \emph{Erd\H{o}s--Gy\'arf\'as
conjecture for cubic planar graphs}, Electron.\ J.\ Combin.\ 20(2) (2013) \#P7.
\bibitem{shauger} S.~E.~Shauger, \emph{Results on the Erd\H{o}s--Gy\'arf\'as
conjecture in $K_{1,m}$-free graphs}, Congressus Numerantium 134 (1998).
\bibitem{daniel-shauger} D.~Daniel, S.~E.~Shauger, \emph{A result on the
Erd\H{o}s--Gy\'arf\'as conjecture in planar graphs}, Congressus Numerantium 153
(2001).
\bibitem{gao-shan} Y.~Gao, S.~Shan, \emph{Erd\H{o}s--Gy\'arf\'as conjecture for
$P_8$-free graphs}, Graphs Combin.\ 38(6) (2022),
DOI 10.1007/s00373-022-02578-9; arXiv:2109.01277.
\bibitem{hu-shen} Z.~Hu, C.~Shen, \emph{The Erd\H{o}s--Gy\'arf\'as conjecture holds
for $P_{10}$-free graphs}, Discrete Math.\ 347(9) (2024) 114175,
DOI 10.1016/j.disc.2024.114175; arXiv:2308.05675.
\bibitem{hegde2024} A.~S.~Hegde, R.~B.~Sandeep, P.~Shashank, \emph{Erd\H{o}s--Gy\'arf\'as
conjecture on graphs without long induced paths} (proves the $P_{13}$-free case
with the aid of a computer search), arXiv:2410.22842 (2024).
\bibitem{carr2026} A.~Carr, \emph{Every minimal counterexample to the
Erd\H{o}s--Gy\'arf\'as conjecture is predominantly cubic}, arXiv:2605.22844 (2026),
preprint.
\bibitem{sms-tocl} M.~Kirchweger, S.~Szeider, \emph{SAT Modulo Symmetries for graph
generation and enumeration}, ACM Trans.\ Comput.\ Logic 25(3) (2024), Art.\ 18,
DOI 10.1145/3670405.
\bibitem{szeider-survey} S.~Szeider, \emph{SAT Modulo Symmetries: A Survey},
CEUR Workshop Proceedings, Vol.\ 4116 (2025).
\bibitem{jooken-survey} J.~Jooken, \emph{Computer-assisted graph theory},
arXiv:2508.20825 (2025).
\bibitem{glasgow} C.~McCreesh, P.~Prosser, J.~Trimble, \emph{The Glasgow Subgraph
Solver: using constraint programming to tackle hard subgraph isomorphism problem
variants}, in: Graph Transformation (ICGT 2020), LNCS 12150, 316--324.
\bibitem{nauty} B.~D.~McKay, A.~Piperno, \emph{Practical graph isomorphism, II},
J.\ Symbolic Comput.\ 60 (2014) 94--112.
\bibitem{wikipedia-eg} Wikipedia, \emph{Erd\H{o}s--Gy\'arf\'as conjecture}
(secondary source for the commonly-cited general $\ge 17$ bound; the primary
record for the general search should be confirmed with Royle/Markstr\"om).
\end{thebibliography}

\end{document}
